---
title: "What the Bootstrap Actually Resamples"
subtitle: "A short note on a method that is easy to run and easy to misread"
author: "Example Author"
date: 2026-09-03
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    theme: cosmo
    fig-width: 7
    fig-height: 4
knitr:
  opts_chunk:
    dev: svg
    message: false
    warning: false
---

## The claim

The bootstrap replaces an unknown sampling distribution with one we can
enumerate. Given a sample $X_1, \dots, X_n$ drawn from an unknown $F$, and a
statistic $T_n = t(X_1, \dots, X_n)$, we would like to know how $T_n$ varies
across repeated samples from $F$. We cannot draw from $F$. So we draw from
$\hat{F}_n$, the empirical distribution that puts mass $1/n$ on each observed
point, and hope that

$$
\sqrt{n}\,(T_n^{*} - T_n) \ \approx\ \sqrt{n}\,(T_n - \theta)
$$

in distribution, where $T_n^{*}$ is computed on a resample. The approximation is
the whole method, and it is an approximation of one distribution by another, not
a way of extracting extra information from a fixed sample.

## A first illustration

Draw one sample, resample it many times, and compare the bootstrap distribution
of the mean against the truth we can compute because we chose the population.

```{r}
#| label: setup
set.seed(20260903)

population <- function(n) rgamma(n, shape = 2, rate = 0.5)
sample_once <- population(60)

boot_means <- replicate(4000, mean(sample(sample_once, replace = TRUE)))
true_means <- replicate(4000, mean(population(60)))

c(
  sample_mean       = mean(sample_once),
  bootstrap_sd      = sd(boot_means),
  true_sd_of_mean   = sd(true_means),
  asymptotic_sd     = sd(sample_once) / sqrt(60)
)
```

The three standard errors agree to about two significant figures. That is the
bootstrap working: it recovered the spread of the sampling distribution without
being told the population.

```{r}
#| label: fig-two-distributions
#| fig-cap: "The bootstrap distribution is centred on the sample mean, not on the truth. Its shape is what carries information; its location does not."
par(mar = c(4, 4, 1, 1))
plot(density(true_means), lwd = 2, col = "#17202a", main = "",
     xlab = "mean", ylab = "density")
lines(density(boot_means), lwd = 2, col = "#3757d5")
abline(v = mean(sample_once), lty = 2, col = "#3757d5")
abline(v = 2 / 0.5, lty = 2, col = "#17202a")
legend("topright", bty = "n", lwd = 2, col = c("#17202a", "#3757d5"),
       legend = c("sampling distribution of the mean", "bootstrap distribution"))
```

Notice the offset. The bootstrap distribution sits over $\bar{X}$, the sample
mean, while the true sampling distribution sits over $\mu$. The bootstrap says
nothing about that gap, which is precisely the bias it cannot see.

## Where it breaks

The approximation relies on $T_n$ being a smooth functional of $F$. The sample
maximum is not. Resampling can never produce a value larger than the largest
observation, so the bootstrap distribution of the maximum is degenerate at the
top and hopeless below it.

```{r}
#| label: fig-maximum
#| fig-cap: "For the sample maximum, the bootstrap distribution is a poor caricature of the truth: it is discrete, bounded above by the observed maximum, and puts about 63% of its mass on that single point."
maxima_true <- replicate(4000, max(population(60)))
maxima_boot <- replicate(4000, max(sample(sample_once, replace = TRUE)))

par(mar = c(4, 4, 1, 1))
hist(maxima_true, breaks = 40, col = "#17202a22", border = NA, freq = FALSE,
     main = "", xlab = "sample maximum", ylim = c(0, 0.45))
hist(maxima_boot, breaks = 40, col = "#3757d555", border = NA, freq = FALSE, add = TRUE)
legend("topright", bty = "n", fill = c("#17202a22", "#3757d555"), border = NA,
       legend = c("truth", "bootstrap"))
```

The mass on the observed maximum is not a numerical accident. A resample misses
a given observation with probability $(1 - 1/n)^n \to e^{-1}$, so the largest
observation appears in about

$$
1 - \left(1 - \tfrac{1}{n}\right)^{n} \longrightarrow 1 - e^{-1} \approx 0.632
$$

of all resamples, and whenever it appears it *is* the maximum.

```{r}
#| label: tbl-mass
#| tbl-cap: "Observed against predicted share of resamples containing the largest observation."
observed <- mean(maxima_boot == max(sample_once))
knitr::kable(
  data.frame(
    quantity  = c("observed share", "1 - (1 - 1/n)^n", "limit 1 - 1/e"),
    value     = round(c(observed, 1 - (1 - 1 / 60)^60, 1 - exp(-1)), 4)
  )
)
```

## Coverage is the honest test

A confidence interval is a claim about repeated use. The way to check one is to
build it many times and count.

```{r}
#| label: fig-coverage
#| fig-cap: "Coverage of a 95% percentile interval for the mean, as the sample grows. The method is asymptotically correct and visibly optimistic when n is small."
coverage <- function(n, reps = 400, B = 400) {
  hits <- replicate(reps, {
    x <- population(n)
    b <- replicate(B, mean(sample(x, replace = TRUE)))
    ci <- quantile(b, c(0.025, 0.975))
    ci[1] <= 4 && 4 <= ci[2]
  })
  mean(hits)
}

sizes <- c(10, 20, 40, 80, 160)
rates <- vapply(sizes, coverage, numeric(1))

par(mar = c(4, 4, 1, 1))
plot(sizes, rates, type = "b", pch = 19, lwd = 2, col = "#3757d5",
     ylim = c(0.8, 1), log = "x", xlab = "n (log scale)", ylab = "coverage")
abline(h = 0.95, lty = 2)
```

At $n = 10$ the nominal 95% interval covers rather less than 95% of the time.
The interval is not wrong so much as over-confident, and no number of bootstrap
replicates $B$ fixes it: $B$ controls Monte Carlo error in the resampling, while
$n$ controls how well $\hat{F}_n$ stands in for $F$.

## What to take away

1. The bootstrap approximates a *distribution*, and inherits any bias in the
   statistic rather than correcting it.
2. It requires smoothness. Extremes, boundaries and non-smooth functionals break
   it in ways that are obvious once plotted and invisible in a table of standard
   errors.
3. $B$ and $n$ are not interchangeable. Raising $B$ buys precision in the
   simulation; only more data buys accuracy in the inference.
