---
title: "Your First Reproducible Report"
format: html
execute:
  echo: true
  warning: false
---

## Make this report yours

This is your private Quarto starter. Edit this paragraph, rename the report,
or share it when you are ready. Your writing is saved without running code.

::: {.callout-note}
### Start with the draft

You can write and read the draft without installing Quarto. The examples below
produce a table and a figure when you render with Quarto locally. This starter
does not contain previously computed results.
:::

## Explore a small dataset

The built-in R dataset `cars` records speed and stopping distance. No data
download is needed. These first six observations give us a starting point:

```{r}
#| label: tbl-observations
#| tbl-cap: "The first six observations in cars."
knitr::kable(head(cars), col.names = c("Speed (mph)", "Stopping distance (ft)"))
```

The scatter plot in @fig-stopping shows how stopping distance varies with speed.
Try changing the point color, then render again to see the change.

```{r}
#| label: fig-stopping
#| fig-cap: "Stopping distance and speed in the cars dataset."
#| fig-alt: "A scatter plot showing stopping distances generally increasing with speed."
plot(cars$speed, cars$dist,
     xlab = "Speed (mph)", ylab = "Stopping distance (ft)",
     pch = 19, col = "steelblue")
```

## Render and keep the results

1. Download the source as `getting-started.qmd` into a local project folder.
2. Install Quarto, R, and the R packages `knitr` and `rmarkdown` on your machine.
3. Link that folder using LibrePaper's local rendering setup. Keep the local
   source synchronized with this document before requesting a render.
4. Choose Render to compute the examples and save their results. Switch between
   the draft and full output to compare your writing with the rendered report.

You can also render the downloaded file directly in a terminal:

```sh
quarto render getting-started.qmd
```

Saved results stay available to readers without your local tools. After editing
code or inputs, check the freshness status and render again when you want new
results. Refresh asks Quarto to recompute; editing prose alone never runs code.

## Next steps

Replace the example data with your analysis, add a caption that explains the
finding, and write a paragraph about what the figure cannot establish. Keep
private data in your local project unless you deliberately choose to share it.
