---
title: "Writing around an R analysis"
format: html
execute:
  cache: true
---

This example is editable as Quarto source. Saved figures stay available while
collaborators revise this paragraph. Execute the document locally to refresh
its analysis.

```{r}
#| label: setup
#| include: false
x <- 1:8
```

```{r}
#| label: fig-first
#| fig-cap: "The original sequence."
plot(x)
```

The second plot uses the same plotting code after changing its input. These
are separate cells and must retain separate results.

```{r}
#| label: change-input
#| include: false
x <- (1:8)^2
```

```{r}
#| label: fig-second
#| fig-cap: "The squared sequence."
plot(x)
```

```{r}
#| label: tbl-summary
#| echo: false
#| tbl-cap: "Summary of the current sequence."
knitr::kable(data.frame(statistic = c("Minimum", "Maximum"), value = range(x)))
```

::: {.callout-note}
Changing either hidden setup cell makes both saved figures potentially outdated.
Editing this explanation keeps the figures available.
:::

```{r}
#| label: documented-only
#| eval: false
stop("This example must never execute")
```
